---
title: "Homework 1: R Fundamentals, Matrix Algebra, and Simulation"
author: "Yunshan Xu"
format: html
---

## Part 1: Vectorization and Data Structures (15 points)

### 1.1 Vectorization (8 points)

Using the built-in `iris` dataset:

a)  (2 points) Create a function `vectorized_zscore` that standardizes a numeric vector using vectorized operations (z = (x - mean(x)) / sd(x)). Create a second function `loop_zscore` that performs the same computation using a `for` loop.

```{r}
# Vectorized version
vectorized_zscore <- function(x) {
  (x - mean(x)) / sd(x)
}

# Loop version
loop_zscore <- function(x) {
  n <- length(x)
  z <- numeric(n)         # pre-allocate 
  m <- mean(x)
  s <- sd(x)
  for (i in seq_len(n)) {
    z[i] <- (x[i] - m) / s
  }
  z
}
```

b)  (3 points) Use `bench::mark()` to compare performance with at least 50 iterations. Create a bar plot comparing median execution times.

```{r}
library(bench)

x_iris <- iris$Sepal.Length

bm <- bench::mark(
  vectorized = vectorized_zscore(x_iris),
  loop       = loop_zscore(x_iris),
  iterations = 50,
  check      = TRUE    
)

# Bar plot of median execution times
medians <- as.numeric(bm$median)         
names(medians) <- c("Vectorized", "Loop")

barplot(
  medians * 1e6,                         # convert to microseconds
  names.arg = names(medians),
  ylab      = "Median execution time (microseconds)",
  main      = "Vectorized vs Loop Z-score: Median Execution Time",
  col       = c("steelblue", "tomato")
)

```

c\) (3 points) Using the \`ChickWeight\` dataset, create a \`Growth.Rate\` column (weight change per day for each chick) and a \`Weight.Gain.Percent\` column (percentage gain from first to last measurement). Calculate average growth rate and mean maximum weight by diet group.

```{r}
library(dplyr)

cw <- ChickWeight %>%
  group_by(Chick) %>%
  arrange(Time) %>%
  mutate(
    # weight change per day between consecutive measurements
    Growth.Rate = (weight - lag(weight)) / (Time - lag(Time)),
    # percentage gain from first to last measurement
    Weight.Gain.Percent = (weight - first(weight)) / first(weight) * 100
  ) %>%
  ungroup()

# Average growth rate by diet group
avg_growth <- cw %>%
  group_by(Diet) %>%
  summarise(Avg.Growth.Rate = mean(Growth.Rate, na.rm = TRUE))

# Mean of each chick's maximum weight, by diet
mean_max_weight <- cw %>%
  group_by(Diet, Chick) %>%
  summarise(Max.Weight = max(weight), .groups = "drop") %>%
  group_by(Diet) %>%
  summarise(Mean.Max.Weight = mean(Max.Weight))

print(avg_growth)
print(mean_max_weight)
```

### 1.2 Data Structures and Functional Programming (7 points)

Using the `palmerpenguins` dataset:

a)  (2 points) Extract the unique species and create a list where each element contains data for one species. Add a sample size attribute to each element.

```{r}
library(palmerpenguins)

species_list <- lapply(unique(penguins$species), function(sp) {
  df <- penguins[penguins$species == sp, ]
  attr(df, "sample_size") <- nrow(df)
  df
})
names(species_list) <- unique(penguins$species)

# Check
attr(species_list[["Adelie"]], "sample_size")

```

b)  (1 point) Use `lapply()` to calculate the mean of each numeric variable (excluding NAs). Redo using `map_dbl()`.

    ```{r}
    library(purrr)

    numeric_cols <- penguins[, sapply(penguins, is.numeric)]

    # lapply version
    lapply(numeric_cols, mean, na.rm = TRUE)

    # map_dbl version (same result, guaranteed numeric output)
    map_dbl(numeric_cols, ~ mean(.x, na.rm = TRUE))
    ```

c)  (1 point) Use `tapply()` to find the mean body mass by species and by the combination of species and sex.

    ```{r}
    # Mean body mass by species
    tapply(penguins$body_mass_g, penguins$species, mean, na.rm = TRUE)

    # Mean body mass by species AND sex combination
    tapply(penguins$body_mass_g,
           list(penguins$species, penguins$sex),
           mean, na.rm = TRUE)
    ```

d)  (1 point) Demonstrate R's copy-on-modify semantics: create a vector, create a reference to it, modify the reference, and explain what happens (2-3 sentences).

    ```{r}
    x <- c(1, 2, 3)
    y <- x         
    y[1] <- 99      

    # x is unchanged
    print(x)  
    print(y)   
    ```

    When y \<- x is assigned, R does not immediately copy the data; both variables point to the same memory location to save space. Only when y is modified does R create a separate copy of the vector, leaving x unchanged. This "copy-on-modify" behavior means R looks memory-efficient until you actually change a shared object, at which point the cost of copying appears.

e)  (2 points) Rewrite the following inefficient code using preallocation. Compare execution times with `system.time()` and explain why your version is faster (2-3 sentences).

```{r}
# Original inefficient version
system.time({
  result <- numeric(0)
  for (i in 1:10000) {
    result <- c(result, i^2)   # copies entire vector each iteration
  }
})

# Pre-allocated version
system.time({
  result_fast <- numeric(10000)  # allocate all memory upfront
  for (i in 1:10000) {
    result_fast[i] <- i^2       
  }
})

```

The growing version is slow because c(result, i\^2) must copy the entire existing vector on every iteration, giving O(n\^2) total work. Pre-allocating with numeric(10000) reserves all memory upfront so each assignment just writes to an existing slot, reducing the work to O(n). As the slides show, this difference becomes dramatic as n grows.

## Part 2: Matrix Operations and Decompositions (30 points)

### 2.1 Matrix Operations (8 points)

Consider the following matrices:

``` r
A <- matrix(c(4, 2, 1, 5), nrow = 2)
B <- matrix(c(3, 1, 0, 2), nrow = 2)
C <- matrix(c(1, 3, 5, 2, 4, 6), nrow = 2)
```

a)  (4 points) Calculate and explain the operation performed in each case: `A + B`, `A * B`, `A %*% B`, `t(A) %*% C`, `solve(A)`.

    ```{r}
    A <- matrix(c(4, 2, 1, 5), nrow = 2)
    B <- matrix(c(3, 1, 0, 2), nrow = 2)
    C <- matrix(c(1, 3, 5, 2, 4, 6), nrow = 2)
    # Element-wise addition: adds corresponding entries
    A + B     

    # Element-wise multiplication: multiplies corresponding entries (NOT matrix mult)
    A * B 

     # True matrix multiplication: dot products of rows of A with columns of B
    A %*% B   

    # Transpose of A (2x2 -> 2x2) then matrix multiply with C (2x3) -> result is 2x3
    t(A) %*% C 

    # Matrix inverse A^{-1}: the matrix such that A %*% solve(A) = I
    solve(A)  

    # Verify solve
    round(A %*% solve(A), 10)  # should be identity matrix

    ```

b)  (4 points) Create a function `is_symmetric` that returns TRUE if a matrix is symmetric and FALSE otherwise. Print an error if the input is not square. Test on A, B, and `t(A) %*% A`.

    ```{r}
    is_symmetric <- function(mat) {
      if (!is.matrix(mat)) stop("Input must be a matrix")
      if (nrow(mat) != ncol(mat)) stop("Matrix must be square")
      all(mat == t(mat))
    }

    is_symmetric(A)           
    is_symmetric(B)           
    is_symmetric(t(A) %*% A)

    ```

### 2.2 Eigenvalue Decomposition (10 points)

a)  (3 points) Compute eigenvalues and eigenvectors using `eigen()`.

    ```{r}
    M <- matrix(c(4, 2, 2, 3), nrow = 2)

    eig_M <- eigen(M)
    lambda <- eig_M$values    # eigenvalues
    P      <- eig_M$vectors   # eigenvectors (columns)

    cat("Eigenvalues:\n"); print(lambda)
    cat("Eigenvectors:\n"); print(P)
    ```

b)  (3 points) Verify the eigenvector equation Mx = lx holds for each pair.

    ```{r}
    for (i in 1:2) {
      x  <- P[, i]
      lx <- lambda[i]
      cat("Eigenpair", i, ":\n")
      cat("  Mx      =", M %*% x, "\n")
      cat("  lambda*x =", lx * x, "\n")
      cat("  Match?  ", isTRUE(all.equal(as.vector(M %*% x), lx * x)), "\n")
    }
    ```

c)  (4 points) Reconstruct M using M = PDP\^{-1} and verify it matches the original.

    ```{r}

    D <- diag(lambda)
    M_reconstructed <- P %*% D %*% solve(P)

    cat("Original M:\n");       print(M)
    cat("Reconstructed M:\n");  print(round(M_reconstructed, 10))
    cat("Match?", isTRUE(all.equal(M, M_reconstructed)), "\n")

    ```

### 2.3 SVD and QR Decomposition (12 points)

a)  **SVD** (6 points): Compute the SVD of the matrix below. Create a rank-1 approximation, calculate the Frobenius norm of the difference, and explain what information is lost.

```{r}
X <- matrix(c(3, 2, 2, 2, 3, -2, 2, -2, 3), nrow = 3)

svd_X <- svd(X)
U <- svd_X$u
D_svd <- diag(svd_X$d)
V <- svd_X$v

cat("Singular values:", svd_X$d, "\n")

# Rank-1 approximation: keep only the first singular value/vector pair
X_rank1 <- svd_X$d[1] * (svd_X$u[, 1] %*% t(svd_X$v[, 1]))

# Frobenius norm of the difference
frob_error <- norm(X - X_rank1, "F")
cat("Frobenius norm of X - X_rank1:", frob_error, "\n")
```

b)  **QR and Least Squares** (6 points): Using the data below, create the design matrix, solve the least squares problem via QR decomposition, compare with `lm()`, and explain the advantages of QR decomposition for least squares.

```{r}
set.seed(228)
n <- 100
x <- runif(n, -5, 5)
y <- 3 + 2 * x + rnorm(n, sd = 1)

# Design matrix with intercept column
X_design <- cbind(1, x)

# QR decomposition approach
qr_X   <- qr(X_design)
Q_mat  <- qr.Q(qr_X)
R_mat  <- qr.R(qr_X)
beta_qr <- solve(R_mat) %*% t(Q_mat) %*% y

# lm() comparison
fit_lm <- lm(y ~ x)

cat("QR coefficients:  ", beta_qr, "\n")
cat("lm() coefficients:", coef(fit_lm), "\n")
cat("Match?", isTRUE(all.equal(as.vector(beta_qr),
                               as.vector(coef(fit_lm)))), "\n")
```

QR is preferred over the normal equations (solve(t(X) %\*% X) %\*% t(X) %\*% y) because forming X'X squares the condition number, amplifying numerical errors. QR preserves the original conditioning of X, making it more stable for ill-conditioned design matrices. R's lm() uses QR internally for exactly this reason.

## Part 3: Optimization and Simulation (40 points)

### 3.1 Gradient Descent (15 points)

Minimize the Rosenbrock function: f(x,y) = (1-x)\^2 + 100(y-x^2)^2

a)  (3 points) Write `rosenbrock(x)` that takes a vector x = c(x1, x2) and returns the function value.

    ```{r}
    rosenbrock <- function(x) {
      (1 - x[1])^2 + 100 * (x[2] - x[1]^2)^2
    }
    ```

b)  (5 points) Write `rosenbrock_grad(x)` that returns the gradient vector.

    ```{r}
    rosenbrock_grad <- function(x) {
      dfdx1 <- -2 * (1 - x[1]) - 400 * x[1] * (x[2] - x[1]^2)
      dfdx2 <- 200 * (x[2] - x[1]^2)
      c(dfdx1, dfdx2)
    }
    ```

c)  (4 points) Implement `gradient_descent(init, grad_fn, learning_rate, max_iter, tol)` that returns the final solution, function value, and iteration count. Stop when the Euclidean norm of the gradient is less than the tolerance.

    ```{r}
    gradient_descent <- function(init, grad_fn, learning_rate, max_iter, tol) {
      x <- init
      iter <- 0

      for (i in 1:max_iter) {
        g <- grad_fn(x)
        
        if (sqrt(sum(g^2)) < tol) {
          iter <- i
          break
        }

        x <- x - learning_rate * g
        iter <- i
      }

      list(
        solution      = x,
        value         = rosenbrock(x),
        iterations    = iter
      )
    }
    ```

d)  (3 points) Apply from starting point (-1.2, 1.0) with learning_rate = 0.001, max_iter = 10000, tol = 1e-6. Report the solution, function value, and iterations.

    ```{r}
    gd_result <- gradient_descent(
      init          = c(-1.2, 1.0),
      grad_fn       = rosenbrock_grad,
      learning_rate = 0.001,
      max_iter      = 10000,
      tol           = 1e-6
    )

    cat("Gradient Descent:\n")
    cat("  Solution:   ", gd_result$solution, "\n")
    cat("  Value:      ", gd_result$value, "\n")
    cat("  Iterations: ", gd_result$iterations, "\n")
    ```

### 3.2 Comparison with Built-in Methods (10 points)

a)  (5 points) Use `optim()` with "BFGS" to minimize the Rosenbrock function from the same starting point.

    ```{r}
    bfgs_result <- optim(
      par    = c(-1.2, 1.0),
      fn     = rosenbrock,
      gr     = rosenbrock_grad,
      method = "BFGS",
      control = list(maxit = 10000)
    )

    cat("\nBFGS:\n")
    cat("  Solution:          ", bfgs_result$par, "\n")
    cat("  Value:             ", bfgs_result$value, "\n")
    cat("  Function evals:    ", bfgs_result$counts["function"], "\n")
    cat("  Gradient evals:    ", bfgs_result$counts["gradient"], "\n")
    ```

b)  (5 points) Use `optim()` with "Nelder-Mead". Compare all three methods (your gradient descent, BFGS, Nelder-Mead) in terms of solution accuracy, iterations, and function evaluations.

    ```{r}
    nm_result <- optim(
      par    = c(-1.2, 1.0),
      fn     = rosenbrock,
      method = "Nelder-Mead",
      control = list(maxit = 10000)
    )

    cat("\nNelder-Mead:\n")
    cat("  Solution:          ", nm_result$par, "\n")
    cat("  Value:             ", nm_result$value, "\n")
    cat("  Function evals:    ", nm_result$counts["function"], "\n")

    ```

    ```{r}
    # Comparison table
    comparison <- data.frame(
      Method      = c("Gradient Descent", "BFGS", "Nelder-Mead"),
      x1          = c(gd_result$solution[1], bfgs_result$par[1], nm_result$par[1]),
      x2          = c(gd_result$solution[2], bfgs_result$par[2], nm_result$par[2]),
      f_value     = c(gd_result$value,       bfgs_result$value,  nm_result$value),
      Iterations  = c(gd_result$iterations,  NA,                 NA),
      Fn_evals    = c(NA, bfgs_result$counts["function"], nm_result$counts["function"])
    )
    print(comparison)
    ```

### 3.3 Simulation Study (15 points)

Compare the mean, median, and 10% trimmed mean as estimators of central tendency.

a)  (7 points) Write `run_simulation(n_samples, n_reps, dist_type)` that generates `n_reps` samples of size `n_samples`, computes all three estimators, and returns a data frame.

    ```{r}
    run_simulation <- function(n_samples, n_reps, dist_type) {
      results <- matrix(NA, nrow = n_reps, ncol = 3,
                        dimnames = list(NULL, c("mean", "median", "trimmed")))

      for (i in 1:n_reps) {
        x <- switch(dist_type,
          "normal" = rnorm(n_samples, 0, 1),
          "t3"     = rt(n_samples, df = 3),
          "contaminated" = {
            n_clean <- rbinom(1, n_samples, 0.9)
            c(rnorm(n_clean, 0, 1), rnorm(n_samples - n_clean, 0, 10))
          }
        )
        results[i, "mean"]    <- mean(x)
        results[i, "median"]  <- median(x)
        results[i, "trimmed"] <- mean(x, trim = 0.1)
      }

      as.data.frame(results)
    }
    ```

b)  (4 points) Run with n = 30, 1000 replications, under three distributions: normal(0,1), t(df=3), and contaminated normal (90% from N(0,1), 10% from N(0,10)). Calculate bias, variance, and MSE for each estimator under each distribution.

    ```{r}
    set.seed(228)
    true_mean <- 0   # all distributions are centered at 0

    distributions <- c("normal", "t3", "contaminated")
    all_metrics <- list()

    for (dist in distributions) {
      sim <- run_simulation(n_samples = 30, n_reps = 1000, dist_type = dist)

      metrics <- data.frame(
        Distribution = dist,
        Estimator    = c("mean", "median", "trimmed"),
        Bias         = colMeans(sim) - true_mean,
        Variance     = apply(sim, 2, var),
        MSE          = colMeans((sim - true_mean)^2)
      )
      all_metrics[[dist]] <- metrics
    }

    summary_table <- do.call(rbind, all_metrics)
    rownames(summary_table) <- NULL
    print(summary_table)
    ```

c)  (4 points) Create a summary table and write a brief paragraph (3-5 sentences) interpreting which estimator performs best under which conditions and why.

    Under the normal distribution, the mean has the lowest MSE because it is the minimum variance unbiased estimator for a symmetric bell-shaped distribution. Under the heavy-tailed t(df=3) distribution, the median and trimmed mean outperform the mean because they are less sensitive to the extreme values that the t-distribution produces frequently. Under contaminated normal data, where 10% of observations come from N(0,10), the trimmed mean tends to perform best overall because it automatically discards the most extreme observations while retaining more data than the median, giving it lower variance than the median and lower sensitivity to outliers than the mean.

------------------------------------------------------------------------

## Part 4: Bootstrap Methods (15 points)

### 4.1 Bootstrap Confidence Intervals (8 points)

a)  (4 points) Implement `bootstrap_mean()` that estimates the standard error and 95% confidence interval for the mean using the nonparametric bootstrap. Support three CI types: "percentile", "basic", and "normal". Apply to the returns data below and report all three intervals.

    ```{r}

    set.seed(42)
    returns <- rnorm(50, mean = 0.08, sd = 0.12)
    returns <- returns + rexp(50, rate = 20) - 0.05

    bootstrap_mean <- function(x, B = 2000, ci_type = "percentile") {
      n          <- length(x)
      obs_mean   <- mean(x)
      boot_means <- numeric(B)

      # resample WITH replacement, same size as original
      for (i in 1:B) {
        boot_sample   <- sample(x, size = n, replace = TRUE)
        boot_means[i] <- mean(boot_sample)
      }

      # SE_boot = sd of bootstrap statistics
      se_boot <- sd(boot_means)

      # Three CI types
      ci <- switch(ci_type,

        # percentile -- take 2.5th and 97.5th percentiles directly
        "percentile" = quantile(boot_means, c(0.025, 0.975)),

        # basic (pivot): reflects the bootstrap distribution around obs_mean
        # lower = 2*obs - upper_percentile, upper = 2*obs - lower_percentile
        "basic" = c(
          2 * obs_mean - quantile(boot_means, 0.975),
          2 * obs_mean - quantile(boot_means, 0.025)
        ),
        
        # normal -- use SE_boot with z critical value (assumes symmetry)
        "normal" = c(
          obs_mean - 1.96 * se_boot,
          obs_mean + 1.96 * se_boot
        )
      )

      list(
        observed_mean  = obs_mean,
        se_boot        = se_boot,
        ci             = ci,
        boot_dist      = boot_means
      )
    }

    set.seed(42)
    res_pct    <- bootstrap_mean(returns, B = 2000, ci_type = "percentile")
    res_basic  <- bootstrap_mean(returns, B = 2000, ci_type = "basic")
    res_normal <- bootstrap_mean(returns, B = 2000, ci_type = "normal")

    cat("Observed mean:     ", res_pct$observed_mean, "\n")
    cat("Bootstrap SE:      ", res_pct$se_boot, "\n\n")
    cat("Percentile CI:     ", res_pct$ci, "\n")
    cat("Basic CI:          ", res_basic$ci, "\n")
    cat("Normal CI:         ", res_normal$ci, "\n")
    ```

<!-- -->

b)  (4 points) Create a histogram of the bootstrap distribution of the mean with vertical lines indicating the three different confidence intervals. Discuss which CI method is most appropriate for this data and why (3-5 sentences).

    ```{r}
    hist(res_pct$boot_dist,
         breaks = 40,
         main   = "Bootstrap Distribution of the Mean",
         xlab   = "Bootstrap Mean",
         col    = "lightblue",
         border = "white")

    # Vertical lines for three CIs
    abline(v = res_pct$ci,    col = "blue",   lty = 2, lwd = 2)
    abline(v = res_basic$ci,  col = "red",    lty = 2, lwd = 2)
    abline(v = res_normal$ci, col = "green",  lty = 2, lwd = 2)
    abline(v = res_pct$observed_mean, col = "black", lwd = 2)

    legend("topright",
           legend = c("Observed mean", "Percentile", "Basic", "Normal"),
           col    = c("black", "blue", "red", "green"),
           lty    = c(1, 2, 2, 2), lwd = 2)

    ```

The returns data is right-skewed due to the added exponential component, making the normal CI least appropriate because it assumes symmetry in the bootstrap distribution. The percentile CI is most appropriate here because, as the slides explain, it automatically accounts for skewness by reading directly from the empirical bootstrap distribution without any normality assumption. The basic CI also handles skewness but via a pivot correction, and tends to give similar results to the percentile method for moderate skewness.

### 4.2 Bootstrap Hypothesis Testing (7 points)

a)  (4 points) Implement `bootstrap_ttest()` that performs a bootstrap test for the difference in means between two groups. Apply to the data below with 2000 bootstrap samples. Create a visualization showing the bootstrap null distribution with the observed difference marked.

    ```{r}
    set.seed(123)
    control   <- rnorm(40, mean = 10,   sd = 2)
    treatment <- rnorm(40, mean = 11.5, sd = 2)

    bootstrap_ttest <- function(x, y, B = 2000) {
      obs_diff <- mean(y) - mean(x)
      x_shifted <- x - mean(x)
      y_shifted <- y - mean(y)
      combined  <- c(x_shifted, y_shifted)

      nx <- length(x)
      ny <- length(y)

      boot_diffs <- numeric(B)

      for (i in 1:B) {
        # Resample under the null (both groups centered at 0)
        boot_x <- sample(combined, size = nx, replace = TRUE)
        boot_y <- sample(combined, size = ny, replace = TRUE)
        boot_diffs[i] <- mean(boot_y) - mean(boot_x)
      }

      # p-value: proportion of bootstrap diffs as extreme as observed
      p_value <- mean(abs(boot_diffs) >= abs(obs_diff))

      list(
        obs_diff   = obs_diff,
        p_value    = p_value,
        boot_dist  = boot_diffs
      )
    }

    set.seed(123)
    boot_test <- bootstrap_ttest(control, treatment, B = 2000)

    cat("Observed difference:", boot_test$obs_diff, "\n")
    cat("Bootstrap p-value:  ", boot_test$p_value, "\n")

    # Standard t-test comparison
    t_result <- t.test(treatment, control)
    cat("t-test p-value:     ", t_result$p.value, "\n")

    # Visualization
    hist(boot_test$boot_dist,
         breaks = 40,
         main   = "Bootstrap Null Distribution of Mean Difference",
         xlab   = "Difference in Means (treatment - control)",
         col    = "lightgrey",
         border = "white")
    abline(v =  boot_test$obs_diff, col = "red", lwd = 2)
    abline(v = -boot_test$obs_diff, col = "red", lwd = 2, lty = 2)
    legend("topright",
           legend = c("Observed diff", "Mirror (two-sided)"),
           col    = "red", lty = c(1, 2), lwd = 2)
    ```

<!-- -->

b)  (3 points) Compare your bootstrap test results with a standard t-test. Discuss any differences and potential advantages of the bootstrap approach (3-5 sentences).

    Both the bootstrap test (p = 0.001) and the t-test (p = 0.0012) reject the null hypothesis and reach the same conclusion that the treatment group has a significantly higher mean than the control group. The two p-values are very close because the data was generated from normal distributions, meaning the t-test's normality assumption is fully satisfied here. The bootstrap's main advantage would appear with non-normal or skewed data, where it does not rely on any distributional assumptions and instead builds the null distribution directly from the data. Additionally, the bootstrap approach generalizes naturally to any test statistic, not just the difference in means, making it more flexible than the t-test for complex statistics where no analytical formula exists.
