---
title: "PHB 228: Homework 2 - Statistical Models, Visualization, and Tools"
author: "Yunshan Xu (A69034781)"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-float: true
    code-fold: false
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

# Load all required packages
library(MASS)
library(car)
library(lme4)
library(ggplot2)
library(dplyr)
library(palmerpenguins)
library(shiny)
library(parallel)
library(microbenchmark)
```

---

# Part 1: Linear and Generalized Linear Models (34 points)

## 1.1 Linear Models and Diagnostics (16 points)

### 1.1a: Linear Model - mpg ~ wt + hp (4 points)

```{r lm-basic}
# Fit linear model predicting mpg from wt and hp
model_1a <- lm(mpg ~ wt + hp, data = mtcars)
summary(model_1a)
```

**Interpretation:**

*TODO: Interpret the coefficients here. Address:*
- *What does the intercept represent?*
- *What is the effect of wt on mpg (holding hp constant)?*
- *What is the effect of hp on mpg (holding wt constant)?*
- *Comment on R-squared and overall model significance.*

### 1.1b: Diagnostic Plots (5 points)

```{r lm-diagnostics}
#| fig-width: 8
#| fig-height: 4

# Residuals vs Fitted
par(mfrow = c(1, 2))
plot(model_1a, which = 1, main = "Residuals vs Fitted")
plot(model_1a, which = 2, main = "Normal Q-Q Plot")
par(mfrow = c(1, 1))
```

**Interpretation:**

*TODO: Interpret each plot:*
- *Residuals vs Fitted: Is there a pattern? Does it suggest non-linearity or heteroscedasticity?*
- *Normal Q-Q: Do residuals follow a normal distribution? Note any deviations.*

### 1.1c: Multicollinearity Check with VIF (3 points)

```{r vif-check}
# Check variance inflation factors
car::vif(model_1a)
```

**Interpretation:**

*TODO: Interpret VIF values. General rule: VIF > 5 suggests moderate multicollinearity, VIF > 10 is severe. What do these values suggest about wt and hp?*

### 1.1d: Interaction Term and Model Comparison (4 points)

```{r lm-interaction}
# Fit model with interaction term
model_1d <- lm(mpg ~ wt * hp, data = mtcars)
summary(model_1d)

# Compare models with ANOVA
anova(model_1a, model_1d)
```

**Model Preference:**

*TODO: Based on the ANOVA p-value, R-squared improvement, and interpretability, which model do you prefer and why?*

---

## 1.2 Logistic Regression (18 points)

```{r biopsy-setup}
# Load and prepare biopsy dataset
data(biopsy, package = "MASS")
biopsy <- na.omit(biopsy)
biopsy$class <- factor(biopsy$class, levels = c("benign", "malignant"))
```

### 1.2a: Boxplot of V1 by Class (4 points)

```{r biopsy-boxplot}
#| fig-width: 6
#| fig-height: 5

# Boxplot comparing V1 between benign and malignant tumors
ggplot(biopsy, aes(x = class, y = V1, fill = class)) +
  geom_boxplot() +
  labs(
    title = "Cell Thickness (V1) by Tumor Class",
    x = "Tumor Class",
    y = "Cell Thickness (V1)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Interpretation:**

*TODO: Compare the distributions. What do the medians, spreads, and outliers suggest about V1 as a predictor of malignancy?*

### 1.2b: Logistic Regression Model (4 points)

```{r logistic-model}
# Fit logistic regression predicting malignancy from V1, V2, V3
model_logit <- glm(class ~ V1 + V2 + V3,
                   data = biopsy,
                   family = binomial)
summary(model_logit)

# Calculate odds ratios
odds_ratios <- exp(coef(model_logit))
odds_ratios
```

**Interpretation:**

*TODO: Interpret the model. For each predictor, what does the odds ratio mean? Which predictors are statistically significant?*

### 1.2c: Confusion Matrix and Performance Metrics (6 points)

```{r confusion-matrix}
# Calculate predicted probabilities
pred_probs <- predict(model_logit, type = "response")

# Apply threshold of 0.5 to get predicted classes
pred_class <- ifelse(pred_probs >= 0.5, "malignant", "benign")
pred_class <- factor(pred_class, levels = c("benign", "malignant"))

# Confusion matrix
conf_matrix <- table(Predicted = pred_class, Actual = biopsy$class)
conf_matrix

# Calculate performance metrics
TP <- conf_matrix["malignant", "malignant"]
TN <- conf_matrix["benign", "benign"]
FP <- conf_matrix["malignant", "benign"]
FN <- conf_matrix["benign", "malignant"]

accuracy    <- (TP + TN) / sum(conf_matrix)
sensitivity <- TP / (TP + FN)
specificity <- TN / (TN + FP)

cat("Accuracy:   ", round(accuracy, 4), "\n")
cat("Sensitivity:", round(sensitivity, 4), "\n")
cat("Specificity:", round(specificity, 4), "\n")
```

**Interpretation:**

*TODO: What does each metric tell us? Is this a good classifier? Discuss any tradeoffs between sensitivity and specificity in a clinical context.*

### 1.2d: Predict New Observation (4 points)

```{r logistic-predict}
# Predict probability of malignancy for new observation
new_obs <- data.frame(V1 = 5, V2 = 4, V3 = 4)
pred_new <- predict(model_logit, newdata = new_obs, type = "response")
cat("Predicted probability of malignancy:", round(pred_new, 4), "\n")
```

**Interpretation:**

*TODO: Interpret this probability in plain language. What would you advise based on this result?*

---

# Part 2: Mixed-Effects Models (22 points)

## 2a: Simulate Dataset (4 points)

```{r mixed-simulate}
library(lme4)
set.seed(123)

n_participants <- 20
n_items        <- 15
n_conditions   <- 2

# TODO: Build simulation with participant random effects, item random effects,
# condition fixed effect, and residual error.
# Structure: each participant completes n_items in each condition

sim_data <- expand.grid(
  participant = factor(1:n_participants),
  item        = factor(1:n_items),
  condition   = factor(c("control", "treatment"))
)

# Participant random intercepts
participant_intercepts <- rnorm(n_participants, mean = 0, sd = 50)

# Item random intercepts
item_intercepts <- rnorm(n_items, mean = 0, sd = 30)

# Fixed effect of condition (treatment slower by 20ms)
condition_effect <- ifelse(sim_data$condition == "treatment", 20, 0)

# Combine effects
sim_data$reaction_time <-
  500 +
  participant_intercepts[sim_data$participant] +
  item_intercepts[sim_data$item] +
  condition_effect +
  rnorm(nrow(sim_data), mean = 0, sd = 25)

head(sim_data)
```

## 2b: Fit and Compare Three Models (5 points)

```{r mixed-models}
# Model 1: Fixed effects only
m1_fixed <- lm(reaction_time ~ condition, data = sim_data)

# Model 2: Random intercepts for participants only
m2_participant <- lmer(reaction_time ~ condition + (1 | participant),
                       data = sim_data, REML = FALSE)

# Model 3: Random intercepts for participants and items
m3_both <- lmer(reaction_time ~ condition + (1 | participant) + (1 | item),
                data = sim_data, REML = FALSE)

# Compare models
AIC(m1_fixed, m2_participant, m3_both)
anova(m2_participant, m3_both)
```

**Comparison:**

*TODO: Based on AIC and likelihood ratio test, which model fits best? Why does accounting for random effects matter here?*

## 2c: Fixed Effects and ICC (5 points)

```{r mixed-fixed-effects}
# Extract fixed effects from best model
fixef(m3_both)

# Calculate ICC for participants
# ICC = variance_participant / (variance_participant + variance_item + residual)
var_components <- as.data.frame(VarCorr(m3_both))
var_participant <- var_components$vcov[var_components$grp == "participant"]
var_item        <- var_components$vcov[var_components$grp == "item"]
var_residual    <- var_components$vcov[var_components$grp == "Residual"]

ICC_participant <- var_participant / (var_participant + var_item + var_residual)
cat("ICC for participants:", round(ICC_participant, 4), "\n")
```

**Interpretation:**

*TODO: Interpret the fixed effects (intercept and condition). What does the ICC value tell us about how much of the variance in reaction time is attributable to between-participant differences?*

## 2d: Random Slopes Model (5 points)

```{r mixed-random-slopes}
# Fit model with random slopes for condition by participant
m4_slopes <- lmer(reaction_time ~ condition + (condition | participant) + (1 | item),
                  data = sim_data, REML = FALSE)

# Likelihood ratio test vs random intercepts model
anova(m3_both, m4_slopes)
```

**Interpretation:**

*TODO: Does adding random slopes significantly improve model fit? What does this mean substantively -- do participants vary in how much the treatment affects their reaction time?*

## 2e: Individual Participant Regression Lines (3 points)

```{r mixed-participant-plot}
#| fig-width: 8
#| fig-height: 6

# Extract predicted values and plot individual lines per participant
sim_data$fitted <- predict(m3_both)

ggplot(sim_data, aes(x = condition, y = fitted,
                     group = participant, color = participant)) +
  geom_line(alpha = 0.6) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(
    title = "Individual Participant Reaction Time by Condition",
    x = "Condition",
    y = "Fitted Reaction Time (ms)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

# Part 3: Advanced Data Visualization (22 points)

## 3.1 Custom Theme (4 points)

```{r custom-theme}
# Custom theme for scientific publication
theme_penguin <- function() {
  theme_minimal() +
    theme(
      # Panel
      panel.background  = element_rect(fill = "white", color = NA),
      panel.grid.major  = element_line(color = "grey90", linewidth = 0.4),
      panel.grid.minor  = element_blank(),
      panel.border      = element_rect(color = "grey70", fill = NA,
                                       linewidth = 0.5),
      # Text
      plot.title        = element_text(size = 14, face = "bold",
                                       margin = margin(b = 8)),
      plot.subtitle     = element_text(size = 11, color = "grey40",
                                       margin = margin(b = 8)),
      axis.title        = element_text(size = 11, face = "bold"),
      axis.text         = element_text(size = 9, color = "grey30"),
      # Legend
      legend.title      = element_text(size = 10, face = "bold"),
      legend.text       = element_text(size = 9),
      legend.background = element_rect(fill = "white", color = "grey85"),
      legend.margin     = margin(4, 6, 4, 6),
      # Strips (facets)
      strip.text        = element_text(size = 10, face = "bold"),
      strip.background  = element_rect(fill = "grey95", color = "grey70")
    )
}
```

## 3.2 Faceted Scatter Plot (6 points)

```{r faceted-plot}
#| fig-width: 10
#| fig-height: 4

# Colorblind-friendly palette
cb_palette <- c("female" = "#0072B2", "male" = "#E69F00")

ggplot(penguins |> filter(!is.na(sex)),
       aes(x = flipper_length_mm, y = bill_length_mm, color = sex)) +
  geom_point(alpha = 0.7, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  facet_wrap(~ species) +
  scale_color_manual(values = cb_palette) +
  labs(
    title    = "Bill Length vs. Flipper Length by Species",
    subtitle = "With linear regression lines and 95% confidence intervals",
    x        = "Flipper Length (mm)",
    y        = "Bill Length (mm)",
    color    = "Sex"
  ) +
  theme_penguin()
```

## 3.3 Distribution Comparison (5 points)

```{r distribution-plot}
#| fig-width: 8
#| fig-height: 5

ggplot(penguins |> filter(!is.na(sex)),
       aes(x = species, y = body_mass_g, fill = sex)) +
  geom_violin(alpha = 0.4, position = position_dodge(0.8)) +
  geom_boxplot(width = 0.2, alpha = 0.8,
               position = position_dodge(0.8),
               outlier.size = 1) +
  scale_fill_manual(values = cb_palette) +
  scale_y_continuous(limits = c(0, NA),
                     labels = scales::comma) +
  labs(
    title = "Body Mass Distribution by Species and Sex",
    x     = "Species",
    y     = "Body Mass (g)",
    fill  = "Sex"
  ) +
  theme_penguin()
```

## 3.4 Complex Multi-Variable Visualization (7 points)

```{r complex-plot}
#| fig-width: 9
#| fig-height: 6

# TODO: Customize further as needed

ggplot(penguins |> filter(!is.na(sex) & !is.na(bill_depth_mm)),
       aes(x = bill_length_mm, y = bill_depth_mm,
           color = species, size = body_mass_g)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7,
              aes(group = species)) +
  scale_color_brewer(palette = "Set1") +
  scale_size_continuous(range = c(1, 6), name = "Body Mass (g)") +
  labs(
    title    = "Bill Morphology and Body Mass Across Penguin Species",
    subtitle = "Point size reflects body mass; lines show species-level trends",
    x        = "Bill Length (mm)",
    y        = "Bill Depth (mm)",
    color    = "Species"
  ) +
  theme_penguin()
```

**Biological Interpretation:**

*TODO: Write 3-5 sentences interpreting the biological patterns. Consider: What does the separation between species suggest about bill morphology? Is there a relationship between bill dimensions and body mass? What might drive these patterns ecologically?*

---

# Part 4: Interactive Visualization with Shiny (12 points)

> Note: Shiny apps cannot run inline in a rendered Quarto HTML file.
> The app code is provided below and can be run with `shiny::runApp()`.

## 4a-c: Shiny App Code

```{r shiny-app}
#| eval: false

library(shiny)
library(ggplot2)

# UI
ui <- fluidPage(
  titlePanel("MTCars Explorer"),

  sidebarLayout(
    sidebarPanel(
      # 4a: Dropdown to select predictor variable
      selectInput(
        inputId  = "predictor",
        label    = "Select Predictor Variable:",
        choices  = c("Weight (wt)"          = "wt",
                     "Horsepower (hp)"      = "hp",
                     "Displacement (disp)"  = "disp",
                     "Rear axle ratio (drat)" = "drat"),
        selected = "wt"
      ),

      # 4c Enhancement: slider to filter by number of cylinders
      # This allows users to explore how the mpg relationship changes
      # for 4, 6, or 8 cylinder subgroups, adding a meaningful
      # stratification layer to the basic scatter plot.
      sliderInput(
        inputId = "cyl_filter",
        label   = "Filter by Cylinders (min):",
        min     = 4, max = 8, value = 4, step = 2
      )
    ),

    mainPanel(
      # 4a: Plot and model summary outputs
      plotOutput("scatter_plot"),
      verbatimTextOutput("model_summary")
    )
  )
)

# Server
server <- function(input, output) {

  # 4b: Reactive expression to fit the linear model
  fitted_model <- reactive({
    filtered_data <- mtcars[mtcars$cyl >= input$cyl_filter, ]
    lm(as.formula(paste("mpg ~", input$predictor)), data = filtered_data)
  })

  # 4b: Render scatter plot with regression line
  output$scatter_plot <- renderPlot({
    filtered_data <- mtcars[mtcars$cyl >= input$cyl_filter, ]
    ggplot(filtered_data, aes_string(x = input$predictor, y = "mpg")) +
      geom_point(size = 2.5, alpha = 0.8, color = "#0072B2") +
      geom_smooth(method = "lm", se = TRUE, color = "#E69F00") +
      labs(
        title = paste("MPG vs.", input$predictor),
        x     = input$predictor,
        y     = "Miles per Gallon (mpg)"
      ) +
      theme_minimal(base_size = 14)
  })

  # 4b: Render model summary
  output$model_summary <- renderPrint({
    summary(fitted_model())
  })
}

# Run the app
shinyApp(ui = ui, server = server)
```

---

# Part 5: Parallel Computing (10 points)

## 5.1a: Core Detection and safe_cores (2 points)

```{r cores-setup}
# Detect number of logical and physical cores
logical_cores  <- parallel::detectCores(logical = TRUE)
physical_cores <- parallel::detectCores(logical = FALSE)

cat("Logical cores: ", logical_cores, "\n")
cat("Physical cores:", physical_cores, "\n")

# Leave one core free for the OS and other processes
safe_cores <- max(1, physical_cores - 1)
cat("safe_cores:    ", safe_cores, "\n")
```

**Choice of safe_cores:**

*TODO: Explain why you left at least one core free. Mention that using all cores can freeze your system or starve the OS of resources.*

## 5.1b: Benchmarking Sequential vs. Parallel (2 points)

```{r benchmark}
# Function that computes mean with a small delay
slow_mean <- function(x) {
  Sys.sleep(0.01)
  mean(x)
}

# Generate 100 random vectors
set.seed(42)
vectors <- lapply(1:100, function(i) rnorm(100))

# Benchmark sequential vs parallel
bm <- microbenchmark::microbenchmark(
  sequential = lapply(vectors, slow_mean),
  parallel   = parallel::mclapply(vectors, slow_mean, mc.cores = safe_cores),
  times = 3  # keep small since each run takes ~1 second
)
print(bm)
```

**Findings:**

*TODO: Compare median times. How much faster is the parallel version? Does the speedup match the number of cores used?*

## 5.2a: Sequential Bootstrap CI (2 points)

```{r bootstrap-sequential}
bootstrap_ci_sequential <- function(data,
                                    n_bootstrap = 1000,
                                    confidence  = 0.95) {
  boot_means <- numeric(n_bootstrap)
  for (i in seq_len(n_bootstrap)) {
    boot_sample  <- sample(data, length(data), replace = TRUE)
    boot_means[i] <- mean(boot_sample)
  }
  alpha <- 1 - confidence
  ci    <- quantile(boot_means, probs = c(alpha / 2, 1 - alpha / 2))
  c(lower = unname(ci[1]), upper = unname(ci[2]))
}
```

## 5.2b: Parallel Bootstrap CI (2 points)

```{r bootstrap-parallel}
bootstrap_ci_parallel <- function(data,
                                  n_bootstrap = 1000,
                                  confidence  = 0.95) {
  boot_means <- unlist(
    parallel::mclapply(seq_len(n_bootstrap), function(i) {
      mean(sample(data, length(data), replace = TRUE))
    }, mc.cores = safe_cores)
  )
  alpha <- 1 - confidence
  ci    <- quantile(boot_means, probs = c(alpha / 2, 1 - alpha / 2))
  c(lower = unname(ci[1]), upper = unname(ci[2]))
}
```

## 5.2c: Scaling Comparison (2 points)

```{r bootstrap-comparison}
# Test on mtcars$mpg with different bootstrap sizes
sizes <- c(500, 1000, 2000)

results <- lapply(sizes, function(n) {
  t_seq <- system.time(
    bootstrap_ci_sequential(mtcars$mpg, n_bootstrap = n)
  )["elapsed"]

  t_par <- system.time(
    bootstrap_ci_parallel(mtcars$mpg, n_bootstrap = n)
  )["elapsed"]

  data.frame(
    n_bootstrap  = n,
    seq_time_s   = round(t_seq, 3),
    par_time_s   = round(t_par, 3),
    speedup      = round(t_seq / t_par, 2)
  )
})

summary_table <- do.call(rbind, results)
print(summary_table)
```

**Findings:**

*TODO: Does the speedup grow with problem size? Is there a point where parallelization overhead makes it not worth it for small n? Comment on your observations.*
